Tidy Data

### 2014-01-17 tidy-data.tex
Manual merge. Closes #7

\begin{enumeriate}

@@

The four verbs can be, and often are, modified by the ``by'' preposition. We often need to perform group-wise aggregation, transformation and subsetting, to find the largest value in each group, to average over replicates and so on. Combining a verb with a ``by'' operator is a concise way to apply that operation to subsets of a data frame. Many \proglang{SAS} {\sc proc}s possess a {\sc by} statement which allows the operation to be performed by group. They are generally input-tidy. Base \proglang{R} possesses a \code{by()} function, which is input-tidy, but, because it produces a list, is not output-tidy. The \code{ddply()} function from the \pkg{plyr} package is a tidy alternative.

@@

To achieve our goal of finding unusual time courses, we'll start by counting the number of deaths in each hour (\code{hod}) for each cause (\code{cod}) with the tidy \code{count} function. Then we remove missing (and hence uninformative for our purpose) values with \code{subset}. This gives Table~\subref*{tbl:counts:1}.

@@

\begin{Verbatim}
hod2 <- count(deaths, c("hod", "cod"))
hod2 <- subset(hod2, !is.na(hod))
\end{Verbatim}

Next we join the dataset to the \code{codes} dataset which gives informative labels for the disease. These two datasets are connected by the \code{cod} variable. This adds a new variable, \code{disease}, shown in Table~\subref*{tbl:counts:2}).

@@

We are interested in points that are high relative to other causes with the same number of deaths, these are the diseases most different from the overall pattern. To find these unusual points, we fit a robust linear model and plot the residuals, as shown in Figure~\ref{fig:devi-resid}. The plot shows an empty region around a residual of 1.5, so somewhat arbitrarily we select those diseases with a residual greater than 1.5. We do that in two steps: first selecting the appropriate rows from \code{devi} (one row per disease), and then finding the matching time course information in the original summary dataset (24 rows per disease).

@@

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.9\textwidth]{case-study/unusual-big}
    \includegraphics[width=0.9\textwidth]{case-study/unusual-sml}
  \caption{Causes of death with unusual time courses. Overall hourly death rate shown in grey. (Top) Causes of death with more than 350 deaths over a year. (Bottom) Causes of death with less than 350 deaths.  Note that different scales are used on the y axis.}
  \label{fig:disease}
\end{figure}

>>>>

\begin{enumerate}

@@

The four verbs can be, and often are, modified by the ``by'' preposition. We often need group-wise aggregates, transformations and subsets, to pick the biggest in each group, to average over replicates and so on. Combining each of the four verbs with a by operator allows them to operate on subsets of a data frame at a time. Many \proglang{SAS} {\sc proc}s possess a {\sc by} statement which allows the operation to be performed by group, and are generally input-tidy. Base \proglang{R} possesses a \code{by()} function, which is input-tidy, but not output-tidy, because it produces a list. The \code{ddply()} function from the \pkg{plyr} package is a tidy alternative.

@@

To achieve our goal of finding unusual temporal patterns, we do the following. First, we count the number of deaths in each hour (\code{hod}) for each cause (\code{cod}) with the tidy \code{count} function.

@@

\begin{Verbatim}
hod2 <- count(deaths, c("hod", "cod"))
\end{Verbatim}

Then we remove missing (and hence uninformative for our purpose) values with \code{subset}.

\begin{Verbatim}
hod2 <- subset(hod2, !is.na(hod))
\end{Verbatim}

This gives Table~\subref*{tbl:counts:1}.

To provide informative labels for disease, we next join the dataset to the \code{codes} dataset . These two datasets are connected by the \code{cod} variable. This adds a new variable, \code{disease}, shown in Table~\subref*{tbl
:counts:2})

@@

We are interested in points that have high $y$-values, relative to their $x$-neighbours. Controlling for the number of deaths, these points represent the diseases which depart the most from the overall pattern.

To find these unusual points, we fit a robust linear model and plot the residuals,  Figure~\ref{fig:devi-resid}. The plot shows an empty region around a residual of 1.5. So somewhat arbitrarily, we'll select those diseases with a residual greater than 1.5. We do this in two steps: first, we select the appropriate rows from \code{devi} (one row per disease), and then we find the matching temporal course information from the original summary dataset (24 rows per disease).

@@

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.9\textwidth]{case-study/unusual-big}
    \includegraphics[width=0.9\textwidth]{case-study/unusual-sml}
  \caption{Causes of death with unusual temporal courses. Overall hourly death rate shown in grey. (Top) Causes of death with more than 350 deaths over a year. (Bottom) Causes of death with fewer than 350 deaths. Note that the y-axes are on different scales.}
  \label{fig:disease}
\end{figure}
